---
title: "Example code for dimensionality reduction and clustering in R"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(umap)
library(Rtsne)
library(dendextend)
library(dbscan)
```

# Exercise

## Setup

- Import the `iris` data set.

```{r}

head(iris)

```

- Separate the matrix of measurements in a new object named `iris_features`.

```{r}
iris_features <- as.matrix(iris[, 1:4])

head(iris_features)
```

# Exercise

## Apply Principal Components Analysis (PCA)

The `prcomp()` function allows you to standardise the data as part of the principal components analysis itself.

- Apply PCA while centering and scaling the matrix of features.

```{r}
pca_iris <- prcomp(iris_features, center = TRUE, scale = TRUE)
pca_iris

#each of these commands gives some extra information
summary(pca_iris)
str(pca_iris)

#
```

- Examine the PCA output.
  Display the loading of each feature on each principal component.

```{r}
pca_iris$rotation[1:4, 1:4]
#this is the principle component values related to each variable. 
```

```{r}
pca_iris$x
# PC values related to each SAMPLE. 

```

- Use the return value of the PCA to create a `data.frame` called `pca_iris_dataframe` that contains the coordinates projected on principal components.

```{r}
pca_iris_dataframe <- as.data.frame(pca_iris$x)

head(pca_iris_dataframe)
pca_iris_dataframe
```

- Visualise the PCA projection using `ggplot2::geom_point()`.

```{r}
ggplot(pca_iris_dataframe, ) +
  geom_point(aes(x = PC1, y = PC2))

```

### Bonus point

- Color data points according to their class label.

- Store the PCA plot as an object named `pca_iris_species`.

```{r}
ggplot(pca_iris_dataframe,) +
  geom_point(aes(x = PC1, y = PC2, color=iris$Species))

#the more reasonable way to do this would be to take the species column from the iris dataframe, and add it to the x part of the pca output dataframe


head(pca_iris_dataframe)
```

```{r}
pca_iris_species <- ggplot(pca_iris_dataframe,    ) +
  
  
pca_iris_species
```

# Exercise

## Variable loading

- Color a scatter plot of PC1 and PC2 by the value of the variable most strongly associated with the first principal component.

What do you observe?

```{r}
#you need to return to the rotation matrix and find the variable measurement with the largest effect on PCA1

#check the rotation matrix
pca_iris$rotation[1:4, 1:4]
#petal length explains the most variation

#need to create a new dataframe. this needs to have a column added for each sample with the petal length. 

#petal lenght: 
petal_length = iris$Petal.Length 

#x values from pca 
pca_iris_dataframe <- as.data.frame(pca_iris$x)

new_df <- cbind(pca_iris_dataframe, petal_length)


#nwo can plot

ggplot(new_df,) +
  geom_point(aes(x = PC1, y = PC2, color=petal_length))



```

```{r}
ggplot(pca_iris_dataframe,    ) +
  
  

```

> Answer:
> 
> 

## Variance explained

- Compute the variance explained by principal components, using information present in the return value of the `prcomp()` function.

```{r}
#this exercise is to help us work out what variance explained is actually calculating
str(pca_iris)

explained_variance_ratio <- pca_iris$sdev^2 / sum(pca_iris$sdev^2)
explained_variance_ratio
```

- Visualise the variance explained by each principal component using `ggplot2::geom_col()`.

```{r}
variance_dataframe <- data.frame(explained_variance_ratio)
variance_dataframe

#the below code needs to be fixed. 
pca_labels <- cbind(variance_dataframe, data.frame(c("PC1", "PC2", "PC3", "PC4")))
pca_labels

ggplot(variance_dataframe) +
  geom_col(aes(x = c("PC1", "PC2", "PC3", "PC4"), y = "explained_variance_ratio"))
  
  

head(variance_dataframe)
```

```{r}
ggplot(variance_dataframe,    ) +
  
  
  
```

# Exercise

## UMAP

- Apply UMAP on the output of the PCA.

```{r}
#PCA output - coordinates of flowers in PCA space. 
pca_iris$x

set.seed(1) # Set a seed for reproducible results
umap_iris <- umap(pca_iris$x)
umap_iris
```

- Inspect the UMAP output.

```{r}
summary(umap_iris) 

head(umap_iris)

str(umap_iris)

```

- Visualise the UMAP projection using `ggplot2::geom_point()`.

```{r}
umap_iris_dataframe <- as.data.frame(umap_iris$layout)

head(umap_iris_dataframe)
```

```{r}
ggplot(umap_iris_dataframe,aes(x = V1, y = V2)) +
  geom_point(aes(colour = petal_length))

#coloured by petal length, just as an example of what can be done
```



### Bonus point

- Color data points according to their class label.

- Store the UMAP plot as an object named `umap_iris_species`.

```{r}
ggplot(umap_iris_dataframe,aes(x = V1, y = V2)) +
  geom_point(aes(colour = iris$Species))
  
```

```{r}
umap_iris_species <- ggplot(umap_iris_dataframe, aes(x = V1, y = V2)) +
  geom_point(aes(colour = iris$Species))
  
umap_iris_species
```

# Exercise

## t-SNE

- Apply t-SNE and inspect the output.

```{r}
set.seed(1) # Set a seed for reproducible results
tsne_iris <- Rtsne(pca_iris$x, pca=FALSE, check_duplicates = FALSE)
str(tsne_iris)
```

- Use the return value of the t-SNE to create a `data.frame` called `tsne_iris_dataframe` that contains the coordinates.

```{r}
tsne_iris_dataframe <- as.data.frame(tsne_iris$Y)

head(tsne_iris_dataframe)
```

- Visualise the t-SNE projection.

```{r}
ggplot(tsne_iris_dataframe, aes(x = V1, y = V2)) +
    geom_point()
  
  
```

### Bonus points

- Color data points according to their class label.

- Store the t-SNE plot as an object named `tsne_iris_species`.

```{r}

head(tsne_iris_dataframe)


ggplot(tsne_iris_dataframe, aes(x = V1, y = V2)) +
    geom_point(aes(colour = iris$Species))
```

```{r}
tsne_iris_species <- ggplot(tsne_iris_dataframe,    ) +
  
  
tsne_iris_species
```

- Combine PCA, UMAP and t-SNE plots in a single figure.

```{r, fig.height=6, fig.width=6}
cowplot::plot_grid(
  
  
  
  
)
```

# Exercise

## Hierarchical clustering

- Perform hierarchical clustering on the `iris_features` data set,
  using the `euclidean` distance and method `ward.D2`.
  Use the functions `dist()` and `hclust()`.

```{r}
# clustering has 3 parts - calculate distance, create tree, cut tree
dist_iris <- dist(iris_features, method = "euclidean")
hclust_iris_ward <- hclust(dist_iris, method = "ward.D2")
hclust_iris_ward
```

- Plot the clustering tree.
  Use the function `plot()`.

```{r}
plot(hclust_iris_ward)
```

How many clusters would you call from a visual inspection of the tree?

> Answer:
> 
> 

- **Bonus point:** Color leaves by known species (use `dendextend`).

```{r}
iris_hclust_dend <- as.dendrogram(hclust_iris_ward)
labels_colors(iris_hclust_dend) <- as.numeric(iris$Species) #Hack to quickly convert a factor into a number which is what it needs to give it a color
plot(iris_hclust_dend)
```

- Cut the tree in 3 clusters and extract the cluster label for each flower.
  Use the function `cutree()`.

```{r}
iris_hclust_ward_labels <- cutree(hclust_iris_ward, k=3)
iris_hclust_ward_labels
#this outputs a unnamed vector.  



#iris_hclust_ward_labels <- cutree(iris_hclust_dend, k=3)
iris_hclust_ward_labels
#this outputs a named vector. the first row is the flower number. the second row tells you which cluster it is in. 
?cutree
```

- Repeat clustering using 3 other agglomeration methods:

  + `complete`
  + `average`
  + `single`

```{r}
# complete
hclust_iris_complete <- hclust(dist_iris, method = "complete")
iris_hclust_complete_labels <- cutree(hclust_iris_complete, k = 3)
iris_hclust_complete_labels
```

```{r}
# average
hclust_iris_average <- hclust(dist_iris, method = "average")
iris_hclust_average_labels <- cutree(hclust_iris_average, k=3)
iris_hclust_average_labels
```

```{r}
# single
hclust_iris_single <- hclust(dist_iris, method = "single")
iris_hclust_single_labels <- cutree(hclust_iris_single, k=3)
iris_hclust_single_labels
```

- Compare clustering results on scatter plots of the data.

```{r}
iris_clusters_dataframe <- iris
iris_clusters_dataframe$hclust_average <- as.factor(iris_hclust_average_labels)
iris_clusters_dataframe$hclust_complete <- as.factor(iris_hclust_complete_labels)
iris_clusters_dataframe$hclust_single <- as.factor(iris_hclust_single_labels)
iris_clusters_dataframe$hclust_ward <- as.factor(iris_hclust_ward_labels)

iris_clusters_dataframe
```

```{r, fig.height=8, fig.width=10}
plot_average <- ggplot(iris_clusters_dataframe, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(aes(colour=hclust_average))
  
plot_complete <- ggplot(iris_clusters_dataframe, aes(x = Petal.Length, y = Petal.Width)) + geom_point(aes(colour=hclust_complete))
  
  
plot_single <- ggplot(iris_clusters_dataframe, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(aes(colour=hclust_single))
  
plot_ward <- ggplot(iris_clusters_dataframe, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(aes(colour=hclust_ward))
  
cowplot::plot_grid(plot_average, plot_complete, plot_single, plot_ward, labels = c('A', 'B', 'C', 'D')
)
```

# Exercise

## dbscan

- Apply `dbscan` to the `iris_features` data set.

```{r}
dbscan_iris <- dbscan(iris_features, eps = 0.5)
dbscan_iris

```

- Visualise the `dbscan` cluster label on a scatter plot of the data.

```{r}
iris_clusters_dataframe$dbscan <- as.factor(   )
head(iris_clusters_dataframe)
```

```{r}
dbscan_plot <- ggplot(iris_clusters_dataframe,    ) +
  
  
dbscan_plot
```

## hdbscan

- Apply `hdbscan` to the `iris_features` data set.

```{r}
hdbscan_iris <- hdbscan(    )
hdbscan_iris
```

- Visualise the `hdbscan` cluster label on a scatter plot of the data.

```{r}
iris_clusters_dataframe$hdbscan <- as.factor(   )
head(iris_clusters_dataframe)
```

```{r}
hdbscan_plot <- ggplot(iris_clusters_dataframe,    ) +
  
  
hdbscan_plot
```

## Bonus point

- Combine the plots of `dbscan` and `hdbscan` into a single plot.

```{r, fig.height=3, fig.width=6}
cowplot::plot_grid(
  
  
  
)
```

# Exercise

## K-means clustering

- Apply $K$-means clustering with $K$ set to 3 clusters.

```{r}
set.seed(1) # Set a seed for reproducible results
kmeans_iris <- kmeans(   )
kmeans_iris
```

- Inspect the output.

```{r}

```

- Extract the cluster labels.

```{r}

```

- Extract the coordinates of the cluster centers.

```{r}

```

- Construct a data frame that combines the `iris` dataset and the cluster label.

```{r}
iris_labelled <- iris
iris_labelled$Kmeans <- as.factor(   )
head(iris_labelled)
```

- Plot the data set as a scatter plot.

  + Color by cluster label.

```{r}
ggplot(iris_labelled,    ) +
  
  
```

### Bonus point

- Add cluster centers as points in the plot.

```{r}
iris_means_centers <- as.data.frame(   )
iris_means_centers$Kmeans <- as.factor(   )
head(iris_means_centers)
```


```{r}
ggplot(iris_labelled,    ) +
  
  
  
```

# Exercise

## Cross-tabulation with ground truth

- Cross-tabulate cluster labels with known labels.

```{r}
table(   )
```

How many observations are mis-classified by $K$-means clustering?

> Answer:
> 
> 
> 
> 
> 

## Elbow plot

- Plot the "total within-cluster sum of squares" for K ranging from 2 to 10.

```{r}

```

```{r}
get_mean_totss_for_k <- function(k, data) {
  kmeans_out <- kmeans(data, k)
  return(kmeans_out$tot.withinss)
}
k_range <- 2:10
kmean_totwithinss <- vapply(   )
kmean_totwithinss
```

```{r}
kmean_totwithinss_dataframe <- data.frame(
  K = ,
  totss = 
)
head(kmean_totwithinss_dataframe)
```

```{r}
ggplot(kmean_totwithinss_dataframe,    ) +
  
  
  
```

Do you agree that 3 is the optimal number of clusters for this data set?

> Answer:
> 
> 
> 
> 

