---
title: "Example code for single-cell droplet quality control"
author: "Kevin Rue-Albrecht"
date: "05/10/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DropletUtils)
library(tidyverse)
library(ggplot2)
library(cowplot)
library(scDblFinder)
library(Seurat)
library(DoubletFinder)
library(celda)
library(scater)
```

# Exercise

## Empty droplets

-   Import the raw matrix into R; use `DropletUtils`. Make a `SingleCellExperiment` object called `sce`.

```{r}
library(DropletUtils)

sce <- read10xCounts(samples = c("pbm5k" = "raw_feature_bc_matrix"), col.names = TRUE) #colnames represent cell barcodes. you can see that the row names are the genes, just by printing the object without having to specify colnames

sce
```

-   Remove droplets without any count at all.

This will massively speed up downstream steps, while removing unquestionable empty droplets.

```{r}

#assay(sce, "counts") is another way of doing sce@counts
#sce@assays$counts
#colSums(sce@assays$counts)

#sce <- sce[, colSums(sce@assays$counts) > 0]

sce <- sce[, colSums(assay(sce, "counts")) > 0]

sce
```

-   Run `DropletUtils::emptyDrops()`.

```{r}
set.seed(1)
out <- DropletUtils::emptyDrops(
          sce, 
          lower = 100, 
          niters = 1000
)
out

out[1:30, ] #need to run in console. nb, if the total integer count is below 100, it does not compute an FDR. 
#FDR is false discovery rate. the is a multiple testing correction. p value is multiplied by number of tests to give FDR. 
```

**How do you read the output? Which droplets would you consider empty?**

> Answer:

-   Remove empty droplets from the data sets. Call the new object `sce_filtered_emptydrops`.

```{r}

positions <- which(out$FDR <0.01) #which would give you the position

sce_filtered_emptydrops <- sce[, positions]

sce_filtered_emptydrops
```

## Bonus point

-   Draw plots comparing empty droplets and other droplets (e.g., library size, mitochondrial content, PCA).

```{r}
#first you need to make a table with the information we need



plot_data <- tibble(library_size = colSums(assay(sce, "counts")), droplet_status = (out$FDR <0.01))
  
plot_data

ggplot(plot_data) +
  geom_violin(aes(x = as.character(droplet_status, levels = c("NA", "TRUE", "FALSE"), y= library_size))) + 
                scale_y_log10()
#i think there was a little issue with the above code. 

#here, use of the scale_y_log10 is great bc it doesn't transform the actual values, it just changes the scale. 


#Code from Jolet where you can re-order the axis and deal with the issue of NA, which is hard to work with 
plot_data <- tibble(library_size = colSums(assay(sce, "counts")), droplet_status = (out$FDR < 0.01))

plot_data$droplet_status[is.na(plot_data$droplet_status)] <- "empty"
  
plot_data$droplet_status <- factor(as.character(plot_data$droplet_status), levels = c("empty", "FALSE", "TRUE"), labels = c("not_tested", "empty_droplet", "cell"))
  
ggplot(plot_data, aes(x = droplet_status, y = library_size)) + geom_violin() + scale_y_log10()

#empty = no test performed, because you decided it is definitely an empty droplet
#true = non-empty droplet. 
#false = is empty droplet 
```

```{r}
is.mt <- 
# table(is.mt)
plot_data <- tibble(
  
  
  
  
)
ggplot(   ) +
  
  
  
```

```{r}

sce <- 
sce <- 
plot_data <- tibble(
  
  
)
ggplot(   ) +
  
  
```

-   Import the filtered matrix for the same cells and compare their set of filtered cells to yours.

```{r}
sce_filtered_cellranger <- 
  
  


all_barcodes <- colnames(sce)

compare_table <- tibble(
  emptydrops = all_barcodes %in%    ,
  cellranger = all_barcodes %in%    ,
)
table(compare_table)
```

# Exercise

## Doublets

Continue using the data set after you excluded empty droplets.

-   Run `scDblFinder::scDblFinder()`.

```{r}
set.seed(1)
sce_scdblfinder <- scDblFinder::scDblFinder(sce_filtered_emptydrops)
table(sce_scdblfinder$scDblFinder.class)
```

```{r}
colData(sce_scdblfinder)
```

**How do you read the output? Which droplets would you consider doublets?**

> Answer:droplets with scDblFinder.class of doublet

-   Remove doublets from the data set. Call the new object `sce_filtered_scdblfinder`.


```{r}
sce_scdblfinder


is.singlet <- sce_scdblfinder$scDblFinder.class == "singlet"
sce_filtered_scdblfinder <- sce_scdblfinder[, is.singlet]

sce_filtered_scdblfinder
```

## Bonus point

-   Draw plots comparing doublets and other droplets. (e.g., library size, mitochondrial content, PCA)

```{r}
plot_data <- tibble(
  barcode = colnames(sce_scdblfinder),
  library_size = colSums(counts(sce_scdblfinder)), 
  droplet_type = sce_scdblfinder$scDblFinder.class
)
ggplot(plot_data) +
  geom_violin(aes(x= droplet_type, y = library_size))

plot_data
  
```

```{r}
is.mt <- 

plot_data <- tibble(
  
  
  
  
)
ggplot(   ) +
  
  
  
```

```{r}
sce_scdblfinder <- 
sce_scdblfinder <- 
plot_data <- tibble(
  
  
)

ggplot(   ) +
  
  
```

-   Compare results with the output of `chris-mcginnis-ucsf/DoubletFinder`

    -   You will need to make a `Seurat` object.

```{r, message=FALSE}
## Pre-process Seurat object (standard)
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
seu <- 
```

```{r, message=FALSE}
## pK Identification (no ground-truth) 

sweep.res.list <- 

sweep.stats <- 
bcmvn <- 
```

```{r}
ggplot() +
  
  
  
```

```{r}
## Homotypic Doublet Proportion Estimate 
homotypic.prop <- 
nExp_poi <- 
nExp_poi.adj <- 
```

```{r, message=FALSE}
## Run DoubletFinder with varying classification stringencies 
seu <- 
```

```{r}
table(   )
```

```{r, message=FALSE}
seu <- 
```

```{r}
table(   )
```

```{r}
doublets_table <- data.frame(
  
  
  
)
table(doublets_table)
```

# Exercise

## Ambient RNA

-   Run `celda::decontX()` (continue using the data after excluding empty droplets and doublets).

```{r, message=FALSE}

assays(sce_filtered_scdblfinder)

set.seed(1)
sce_decontX <- celda::decontX(sce_filtered_scdblfinder, assayName = "counts")


```

-   Visualize the UMAP layout created by `decontX`; use `plotDimReduceCluster()`.


```{r}
sce_decontX 

umap <- plotDimReduceCluster(x = sce_decontX$decontX_clusters, #this is an integer vector of cell cluster
  dim1 = reducedDim(sce_decontX, "decontX_UMAP")[, 1],
  dim2 = reducedDim(sce_decontX, "decontX_UMAP")[, 2],
)

umap
```

-   Visualize the percentage of contamination estimated by `decontX` in each cell. Use the `plotDecontXContamination()` function.

```{r}
plotDecontXContamination(sce_decontX)
#at this point there was a discussion - should we substract the ambient RNA from all droplets, or just remove the droplets that contain over x% ambient RNA. 
```

**Does any cluster display higher (estimated) levels of ambient RNA?**

> Answer:

-   Visualize the log-normalised expression of marker genes on a UMAP layout faceted by gene; use `plotDimReduceFeature()`.

**Note:** For instance, use `CD3D`, `CD3E`, `GNLY`, `LYZ`, `S100A8`, `S100A9`, `CD79A`, `CD79B`, and `MS4A1`.

<!--
Source: https://bioconductor.org/packages/devel/bioc/vignettes/celda/inst/doc/decontX.pdf
(Section 5.3)
We will use:
- CD3D and CD3E for T-cells,
- LYZ, S100A8, and S100A9 for monocytes,
- CD79A, CD79B, and MS4A1 for B-cells,
- GNLY for NK-cells,
- and PPBP for megakaryocytes
-->

```{r}



rowData(sce_decontX)
rownames(sce_decontX) <- 

rownames(sce_decontX) <- uniquifyFeatureNames(rowData(sce_decontX)$ID, rowData(sce_decontX)$Symbol)

P1 <- plotDimReduceFeature(decontXcounts(sce_decontX), features = c("CD3D", "CD3E", "GNLY", "LYZ", "S100A8", "S100A9", "CD79A", "CD79B", "MS4A1"), 
  dim1 = reducedDim(sce_decontX, "decontX_UMAP")[, 1],
  dim2 = reducedDim(sce_decontX, "decontX_UMAP")[, 2],
)

#COMPARE THE ABOVE WITH:
P2 <- plotDimReduceFeature(counts(sce_decontX), features = c("CD3D", "CD3E", "GNLY", "LYZ", "S100A8", "S100A9", "CD79A", "CD79B", "MS4A1"), 
  dim1 = reducedDim(sce_decontX, "decontX_UMAP")[, 1],
  dim2 = reducedDim(sce_decontX, "decontX_UMAP")[, 2],
)

cowplot::plot_grid(P1, P2)
```

-   Visualise the percentage of cells with detectable expression of marker genes before and after decontamination as a bar plot; use `plotDecontXMarkerPercentage()`.

```{r}

#making markers for the plotDecontXMarkerPercentage fx
markers <- list(
  killer = c('CD3D', 'CD3E'), 
  Bcell = c('LYZ', 'S100A8', 'S100A9'),
  Monocytes = c("CD79A", "CD79B", "MS4A1"), 
  NK_cells = c('GNLY'), 
  Megakaryocytes = c('PPBP'))

  
cellTypeMappings <- list(
  Tcell = c(2), 
  Bcell = c(3),
  Monocytes = c(4), 
  NK_cells = c(1), 
  Megakaryocytes = c(5))

plotDecontXMarkerPercentage(
  sce_decontX, 
  markers = markers,
  groupClusters = cellTypeMappings,
  assayName = c("counts", "decontXcounts"),
)


#labels in grey banner is marker. the x labels are the clusters we've defined. 
```

```{r}
plotDecontXMarkerPercentage(
  sce_decontX,
  
  
  
)
```

```{r}
plotDecontXMarkerPercentage(
  
  
  
  
)
```

-   Visualize the log-normalised expression of marker genes before and after decontamination as a violin plot; use `plotDecontXMarkerExpression()`.

```{r, fig.width=10, fig.height=10}
plotDecontXMarkerExpression(sce_decontX, markers = markers, groupClusters = cellTypeMappings, assayName = c("counts", "decontXcounts")
)

#NB, the decontx counts are just counts, not log normalised counts, which are referred to in an assay called logcounts

#deconXcounts is a new matrix with contaminants removed. so you would do a lognorm counts on this matrix and continue with the workflow that we did in the past. 
#log norm counts by default calls the output logcounts, so you need to specify the name of the output when you transform them

sce_decontX <- logNormCounts(sce_decontX, assay.type = "decontXcounts", name = "log_normalised_decontX_counts")
assays(sce_decontX)

#also normalise the raw counts
sce_decontX <- logNormCounts(sce_decontX, assay.type = "counts", name = "log_normalised_counts")
assays
#plotting again with log normalised count (which is helpful bc there can be very low counts for many genes)
plotDecontXMarkerExpression(sce_decontX, markers = markers, groupClusters = cellTypeMappings, assayName = c("log_normalised_counts", "log_normalised_decontX_counts"))
```
