---
title: "Example code for ggplot2"
author: "Kevin Rue-Albrecht"
date: "03/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(patchwork)
library(gridExtra)
```

# Exercise

Use `library(ggplot2)` to load the package.

```{r}
library(ggplot2)
```

## Specify the dataset.

What does `ggplot(diamonds)` do?

```{r}
ggplot(diamonds)
```

> Answer:

## Add the aesthetics.

What does `ggplot(diamonds, aes(x = carat, y = price))` do?

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

> Answer:

## Add geometric objects

- Add data points showing `carat` on the x-axis and `price` on the y-axis.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) + 
  geom_point()

```

- Color data points by `cut`.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price, colour=cut)) + 
  geom_point()

  
```

- Add a smoothed mean trend line.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(diamonds, aes(x = carat, y = price, colour=cut)) + 
  geom_point() +
  geom_smooth()


```

- Assign that last plot to an object called `obds_diamonds`.

```{r}

obds_diamonds <- ggplot(diamonds, aes(x = carat, y = price, colour=cut)) + 
  geom_point() +
  geom_smooth()


```


# Exercise

Predict the difference between these two plots

### Plot 1

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price, colour = cut)) +
  geom_point() +
  geom_smooth()
```

### Plot 2

```{r, include=TRUE}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(colour = cut)) +
  geom_smooth()
```

> Answer:

# Exercise

## Trend lines

Using the `ChickWeight` data set:

- Create a scatter plot of weight (y-axis) over time (x-axis).

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight)) +
  geom_point()

```

- Color by diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight)) +
  geom_point(aes(colour = Diet)) 


```

- Add a linear mean trend line for each diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight)) +
  geom_point(aes(colour = Diet)) +
  geom_smooth(method = lm)

#this needs to be chaned to the below in order to get a trend line by diet 

ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth(method = lm)

```

At this point you should be able to visually identify
the diet that leads to the highest mean increase in weight.

> Answer:

- Facet a ribbon of sub-plots, one per diet.
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth() + 
  facet_wrap(~Diet)
```

- Assign that last plot to an object called `obds_chickweight`.

```{r}

obds_chickweight <- ggplot(ChickWeight, aes(x = Time, y = weight, colour = Diet)) +
  geom_point() +
  geom_smooth() + 
  facet_wrap(~Diet) 


```

# Exercise

## Bar plot

- Load the `ggplot2::msleep` data set.

```{r}

?ggplot2::msleep
msleep

```

- Draw a bar plot of number of observations (i.e., rows) for each taxonomic order
  (i.e, one plot and one bar per taxonomic order).

```{r}
ggplot(msleep, aes(x=order)) +
  geom_bar()

```

- Change the angle and font size of the text for the x-axis ticks (not the axis titles).
  Justify the text of those x-axis ticks as right-aligned.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x=order)) +
  geom_bar() +
  theme(axis.text.x = element_text(size = rel(0.8), angle = 90, hjust = 1))


```

- Change the value and font size of the title for both x and y axes.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x=order)) +
  geom_bar() +
  theme(axis.text.x = element_text(size = rel(0.8), angle = 90, hjust = 1)) + 
  theme(axis.title = element_text(size = 12))

```

- Fill each bar with colors, proportionally to the count of each genus.
  (Copy-paste and extend the code chunk above.)

```{r}

ggplot(msleep, aes(x=order,fill=genus)) +
  geom_bar() +
  theme(axis.text.x = element_text(size = rel(0.8), angle = 90, hjust = 1)) + 
  theme(axis.title = element_text(size = 12)) 







```

From this point onwards, you may need to iteratively resize the text of the ticks and axes for readability.

- Reduce the legend key size.
  (Recommendation: use `unit(2, "mm")`).
  (Copy-paste and extend the code chunk above.)

```{r}
ggplot(msleep, aes(x=order,fill=genus)) +
  geom_bar() +
  theme(axis.text.x = element_text(size = rel(0.8), angle = 90, hjust = 1)) + 
  theme(axis.title = element_text(size = 12)) + 
  theme(legend.key.size = unit(1, "mm"))





```

- Force the legend to be display in 3 columns.
  (Recommendation: use `guide_legend(...)`).
  (Copy-paste and extend the code chunk above.)

```{r, fig.height=3}

ggplot(msleep, aes(x=order,fill=genus)) +
  geom_bar() +
  theme(axis.text.x = element_text(size = rel(0.8), angle = 90, hjust = 1)) + 
  theme(axis.title = element_text(size = 12)) + 
  theme(legend.key.size = unit(1, "mm")) + 
  guides(fill=guide_legend(ncol=3)) #asyou have filled by genus, you select fillguide

```

- Add a contour of thin black lines to the bars.

```{r}

ggplot(msleep, aes(x=order,fill=genus)) +
  geom_bar(colour='black', linewidth = 0.1) +
  theme(axis.text.x = element_text(size = rel(0.8), angle = 90, hjust = 1)) + 
  theme(axis.title = element_text(size = 12)) + 
  theme(legend.key.size = unit(1, "mm")) + 
  guides(fill=guide_legend(ncol=3)) #asyou have filled by genus, you select fillguide









```

- Assign that last plot to an object called `obds_msleep`.

```{r}


obds_msleep <- ggplot(msleep, aes(x=order,fill=genus)) +
  geom_bar(colour='black', linewidth = 0.1) +
  theme(axis.text.x = element_text(size = rel(0.8), angle = 90, hjust = 1)) + 
  theme(axis.title = element_text(size = 12)) + 
  theme(legend.key.size = unit(1, "mm")) + 
  guides(fill=guide_legend(ncol=3)) #asyou have filled by genus, you select fillguide









```


# Exercise

## Plotting grid

- Collate the plots that we assigend to objects through the day, as a single plot.

  + Plots: `obds_diamonds`, `obds_chickweight`, `obds_msleep`.
  
  + Methods: `cowplot::plot_grid()`, `patchwork`, `gridExtra::grid.arrange()`.

Using `cowplot`.

```{r, fig.height=12, fig.width=12}
library(cowplot)
first_row <- cowplot::plot_grid(obds_chickweight, obds_diamonds)
second_row <- cowplot::plot_grid(obds_msleep)
super_plot <- cowplot::plot_grid(first_row, second_row, nrow = 2)
super_plot
```

```{r, fig.height=12, fig.width=12}
library(cowplot)
plot_grid(obds_chickweight, obds_diamonds, obds_msleep, labels = c("A", "B", "C"), rel_widths = c(1,1,5))

```





Using `patchwork`.

```{r, fig.height=12, fig.width=12}
library(patchwork)

```

Using `gridExtra`.

```{r, fig.height=12, fig.width=12}
library(gridExtra)





```

- Export the new plot in a PDF file, and open it in a PDF viewer (e.g. Adobe Acrobat Reader DC).

You will likely need a few attempts to fine-tune the width and height of the output file.

```{r}
ggsave(   )
```

# Exercise

## Pair programming

- Explore the data set `ggplot2::mpg` and generate the most informative plot that you can!

```{r}




```
